---
title: "Wine Quality Predictor"
output: html_notebook
---

# Dependencies

```{r}
library(dplyr)
library(knitr)
library(tibble)
library(CORElearn)
library(ggplot2)
library(corrplot)
library(rpart)
library(rpart.plot)
library(caret)
library(e1071)
```


# Introduction


# Reading Dataset


**White Wine dataset**
```{r}
whiteWineURL <- "https://archive.ics.uci.edu/ml/machine-learning-databases/wine-quality/winequality-white.csv"

whiteRawData <- read.csv(whiteWineURL, header = TRUE, sep = ";")
```

**Red Wine dataset**

```{r}
redWineURL <- "https://archive.ics.uci.edu/ml/machine-learning-databases/wine-quality/winequality-red.csv"

redRawData <- read.csv(redWineURL, header = TRUE, sep = ";")
```

Add Color attribute

```{r}
redRawData['color'] <- 'red.wine'
whiteRawData['color'] <- 'white.wine'
```


Number of samples in raw data
```{r}
nrow(redRawData)
```


```{r}
nrow(whiteRawData)
```

## Unknown Values ##

Datasets have no unknown values
```{r}
any(is.na(whiteRawData))
any(is.na(redRawData))
```


## Duplicated Samples ##


There are duplicated samples in both of the datasets

```{r}
any(duplicated(whiteRawData))
```

```{r}
any(duplicated(redRawData))
```

Remove duplicated data
```{r}
whiteRawData <- whiteRawData[!duplicated(whiteRawData),]
redRawData <- redRawData[!duplicated(redRawData),]
```

Number of samples are reduced
```{r}
nrow(redRawData)
```


```{r}
nrow(whiteRawData)
```

## Attribute Overview ##

Quality value ranges from 3 to 9 in white wine dataset.
```{r}
unique(whiteRawData$quality)
```


On the other hand quality in red wine dataset ranges from 3 to 8.
```{r}
unique(redRawData$quality)
```


If we look at any dataset, quality values are integer.
This causes trouble in R functions when evaluating quality and other attributes together.

```{r}
kable(sapply(whiteRawData, class))
```


We can cast quality values to numeric.

```{r}
whiteRawData$quality <- as.numeric(whiteRawData$quality)
redRawData$quality <- as.numeric(redRawData$quality)
```

```{r}
mergedRawData <- rbind(whiteRawData, redRawData)
```

Quality values can be categorized as "Poor", "Normal" and "Good".

```{r}
whiteData <- whiteRawData
whiteData$quality <- cut(whiteData$quality, 
                          breaks=c(3, 5, 6, Inf), 
                          labels=c("Poor", "Normal", "Good"), 
                          include.lowest = TRUE)
```


```{r}
redData <- redRawData
redData$quality <- cut(redData$quality, 
                      breaks=c(3, 5, 6, Inf), 
                      labels=c("Poor", "Normal", "Good"), 
                      include.lowest = TRUE)
```


Merged dataset

```{r}
mergedData <- rbind(whiteData, redData)
```

# Explotary Data Analysis


## Custom Functions ## 

```{r}
histogramPlot <- function(data, attribute, binWidth, xLabel, title) 
{
   ggplot(data, aes(x=attribute, fill = quality)) + 
    geom_histogram(binwidth = binWidth, color="black", fill="steelblue", linetype="dashed") +
    ggtitle(title) +
    labs(x = xLabel, y = "Count") + 
    facet_wrap( ~quality) + 
    theme(plot.title = element_text(hjust = 0.5))
}
```

```{r}
histogramByLogScalePlot <- function(data, attribute, binWidth, xLabel, title) 
{
  ggplot(data, aes(x=attribute)) + 
    geom_histogram(binwidth = binWidth, color="black", fill="steelblue", linetype="dashed") +
    scale_x_log10()+
    ggtitle(title) +
    labs(x = xLabel, y = "Count") + 
    facet_wrap( ~color) + 
    theme(plot.title = element_text(hjust = 0.5))
  
}
```

```{r}
attibutesOverQualityBoxplot <- function (attribute, title) 
{
  ggplot(mergedData, aes(x = factor(quality), y = attribute)) + 
    stat_boxplot(geom ='errorbar') + 
    labs(x = "Quality", y = "Value") + 
    ggtitle(title) +
    geom_boxplot(outlier.colour = "red", outlier.shape = 1,  colour = "blue") +
    geom_jitter(size = 1.5, alpha = 0.1)  +
    theme(plot.title = element_text(hjust = 0.5)) +
    facet_wrap(~color)
}

```


## Quality ##

Number of Samples according to numeric qualities.

```{r}
ggplot(data = mergedRawData, aes(x = quality)) +
  geom_bar(fill="steelblue", width=0.5) +
  ggtitle("Number Of Samples According to Quality") +
  labs(x = "Quality", y = "Count") + 
  geom_text(stat = 'count',aes(label =..count.., vjust = -0.2)) +
  facet_wrap( ~color) +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_x_continuous(breaks = unique(mergedRawData$quality))
```

Number of Samples according to categorical qualities.
```{r}
ggplot(data = mergedData, aes(x = quality)) +
  geom_bar(fill="steelblue", width=0.5) +
  ggtitle("Number Of Samples According to Quality") +
  labs(x = "Quality", y = "Count") + 
  geom_text(stat = 'count',aes(label =..count.., vjust = -0.2)) +
  facet_wrap( ~color) +
  theme(plot.title = element_text(hjust = 0.5))
```



## Fixed Acidity ##


```{r}
histogramPlot(whiteData, whiteData$fixed.acidity, 0.5, 
              "Fixed Acidity", "Histogram of White Wine Fixed Acidity")
```

```{r}
histogramPlot(redData, redData$fixed.acidity, 0.5, 
              "Fixed Acidity", "Red Wine Histogram of Red Wine Fixed Acidity")
```

```{r}
attibutesOverQualityBoxplot(mergedData$fixed.acidity, "Fixed Acidity Boxplot")
```

```{r}
redData %>% 
  select(fixed.acidity) %>% 
  summarise_all(list(md = median, mn = mean, sd = sd, IQR = IQR)) %>% kable
```

```{r}
whiteData %>% 
  select(fixed.acidity) %>% 
  summarise_all(list(md = median, mn = mean, sd = sd, IQR = IQR)) %>% kable
```


## Volatile Acidity ##

```{r}
histogramPlot(whiteData, whiteData$volatile.acidity, 0.05, 
              "Volatile Acidity", "Histogram of White Wine Volatile Acidity")
```


```{r}
histogramPlot(redData, redData$volatile.acidity, 0.05, 
              "Volatile Acidity", "Histogram of Red Wine Volatile Acidity")
```


```{r}
attibutesOverQualityBoxplot(mergedData$volatile.acidity, "Volatile Acidity Boxplot")
```


```{r}
redData %>% 
  select(volatile.acidity) %>% 
  summarise_all(list(md = median, mn = mean, sd = sd, IQR = IQR)) %>% kable
```

```{r}
whiteData %>% 
  select(volatile.acidity) %>% 
  summarise_all(list(md = median, mn = mean, sd = sd, IQR = IQR)) %>% kable
```


## Citric Acid  ##
```{r}
histogramPlot(whiteData, whiteData$citric.acid, 0.05, "Citric Acidity", "Histogram of Citric Acidity")
```

```{r}
histogramPlot(redData, redData$citric.acid, 0.05, "Citric Acidity", "Histogram of Citric Acidity")
```


```{r}
attibutesOverQualityBoxplot(mergedData$citric.acid, "Citric Acidity Boxplot")
```



```{r}
redData %>% 
  select(citric.acid) %>% 
  summarise_all(list(md = median, mn = mean, sd = sd, IQR = IQR)) %>% kable
```

```{r}
whiteData %>% 
  select(citric.acid) %>% 
  summarise_all(list(md = median, mn = mean, sd = sd, IQR = IQR)) %>% kable
```


## Residual Sugar  ##

```{r}
histogramPlot(whiteData, whiteData$residual.sugar, 2.5, "Residual Sugar","Histogram of Residual Sugar")
```


```{r}
histogramPlot(redData, redData$residual.sugar, 2.5, "Residual Sugar","Histogram of Residual Sugar")
```


```{r}
attibutesOverQualityBoxplot(mergedData$residual.sugar, "Residual Sugar Boxplot")
```


```{r}
redData %>% 
  select(residual.sugar) %>% 
  summarise_all(list(md = median, mn = mean, sd = sd, IQR = IQR)) %>% kable
```

```{r}
whiteData %>% 
  select(residual.sugar) %>% 
  summarise_all(list(md = median, mn = mean, sd = sd, IQR = IQR)) %>% kable
```

## Chlorides  ##

```{r}
histogramPlot(whiteData, whiteData$chlorides, 0.025, 
              "Chlorides", "Histogram of White Wine Chlorides")
```

```{r}
histogramPlot(redData, redData$chlorides, 0.025, 
              "Chlorides", "Histogram of Red Wine Chlorides")
```


```{r}
attibutesOverQualityBoxplot(mergedData$chlorides, "Chlorides Boxplot")
```


```{r}
redData %>% 
  select(chlorides) %>% 
  summarise_all(list(md = median, mn = mean, sd = sd, IQR = IQR)) %>% kable
```

```{r}
whiteData %>% 
  select(chlorides) %>% 
  summarise_all(list(md = median, mn = mean, sd = sd, IQR = IQR)) %>% kable
```




## Free Sulfur Dioxide##

```{r}
histogramPlot(whiteData, whiteData$free.sulfur.dioxide, 10.0, 
              "Free Sulfur Dioxide","Histogram of White Wine Free Sulfur Dioxide")
```


```{r}
histogramPlot(redData, redData$free.sulfur.dioxide, 5.0, 
              "Free Sulfur Dioxide","Histogram of Red Wine Free Sulfur Dioxide")
```

```{r}
attibutesOverQualityBoxplot(mergedData$free.sulfur.dioxide, "Free Sulfur Dioxide Boxplot")
```



```{r}
redData %>% 
  select(free.sulfur.dioxide) %>% 
  summarise_all(list(md = median, mn = mean, sd = sd, IQR = IQR)) %>% kable
```

```{r}
whiteData %>% 
  select(free.sulfur.dioxide) %>% 
  summarise_all(list(md = median, mn = mean, sd = sd, IQR = IQR)) %>% kable
```




## Total Sulfur Dioxide ##

```{r}
histogramPlot(mergedData, mergedData$total.sulfur.dioxide, 10.0, 
              "Total Sulfur Dioxide", "Histogram of White Wine Total Sulfur Dioxide")
```


```{r}
histogramPlot(redData, redData$total.sulfur.dioxide, 10.0, 
              "Total Sulfur Dioxide", "Histogram of Red Wine Total Sulfur Dioxide")
```

```{r}
attibutesOverQualityBoxplot(mergedData$total.sulfur.dioxide, "Total Sulfur Dioxide Boxplot")
```


```{r}
redData %>% 
  select(total.sulfur.dioxide) %>% 
  summarise_all(list(md = median, mn = mean, sd = sd, IQR = IQR)) %>% kable
```

```{r}
whiteData %>% 
  select(total.sulfur.dioxide) %>% 
  summarise_all(list(md = median, mn = mean, sd = sd, IQR = IQR)) %>% kable
```


## Density ##

```{r}
histogramPlot(whiteData, whiteData$density, 0.001, 
              "Density", "Histogram of White Wine Density")
```



```{r}
histogramPlot(redData, redData$density, 0.001, 
              "Density", "Histogram of Red Wine Density")
```



```{r}
attibutesOverQualityBoxplot(mergedData$density, "Density Boxplot")
```



```{r}
redData %>% 
  select(density) %>% 
  summarise_all(list(md = median, mn = mean, sd = sd, IQR = IQR)) %>% kable
```

```{r}
whiteData %>% 
  select(density) %>% 
  summarise_all(list(md = median, mn = mean, sd = sd, IQR = IQR)) %>% kable
```



## pH ##

```{r}
histogramPlot(whiteData, whiteData$pH, 0.05, "pH","Histogram of White Wine pH")
```


```{r}
histogramPlot(redData, redData$pH, 0.05, "pH","Histogram of Red Wine pH")
```

```{r}
attibutesOverQualityBoxplot(mergedData$pH, "pH Boxplot")
```



```{r}
redData %>% 
  select(pH) %>% 
  summarise_all(list(md = median, mn = mean, sd = sd, IQR = IQR)) %>% kable
```

```{r}
whiteData %>% 
  select(pH) %>% 
  summarise_all(list(md = median, mn = mean, sd = sd, IQR = IQR)) %>% kable
```



## Sulphates ##

```{r}
histogramPlot(whiteData, whiteData$sulphates, 0.05, 
              "Sulphates","Histogram of White Wine Sulphates")
```


```{r}
histogramPlot(redData, redData$sulphates, 0.05, 
              "Sulphates","Histogram of Red Wine Sulphates")
```


```{r}
attibutesOverQualityBoxplot(mergedData$sulphates, "Sulphates Boxplot")
```

```{r}
redData %>% 
  select(sulphates) %>% 
  summarise_all(list(md = median, mn = mean, sd = sd, IQR = IQR)) %>% kable
```

```{r}
whiteData %>% 
  select(sulphates) %>% 
  summarise_all(list(md = median, mn = mean, sd = sd, IQR = IQR)) %>% kable
```



## Alcohol ##

```{r}
histogramPlot(whiteData, whiteData$alcohol, 0.5, 
              "Alcohol","Histogram of White Wine Alcohol")
```


```{r}
histogramPlot(redData, redData$alcohol, 0.5, 
              "Alcohol","Histogram of Red Wine Alcohol")
```


```{r}
attibutesOverQualityBoxplot(mergedData$alcohol, "Alcohol Boxplot")
```

```{r}
redData %>% 
  select(alcohol) %>% 
  summarise_all(list(md = median, mn = mean, sd = sd, IQR = IQR)) %>% kable
```

```{r}
whiteData %>% 
  select(alcohol) %>% 
  summarise_all(list(md = median, mn = mean, sd = sd, IQR = IQR)) %>% kable
```

## Correlation Matrix ##


### Red Wine Dataset ###


```{r}

cex.before <- par("cex")
par(cex = 0.7)
corrplot(as.matrix(cor(subset(redRawData, select=-c(color)))), 
         cl.lim = c(-100, 100), type = "upper", insig = "blank", method = "color",
    addCoef.col="grey", 
    order = "AOE", tl.cex = 1/par("cex"),
    cl.cex = 1/par("cex"), addCoefasPercent = TRUE)
par(cex = cex.before)

```


Quality is correlated positively with Alcohol and Sulphate, negatively with Volatile Acidity.

Also Information Gain supports correlation matrix and gives same attributes.



```{r}
attrEval(quality ~ ., data=select(redData, -color),  estimator = "InfGain") %>% sort(decreasing = TRUE) %>% kable 
```




#```{r}
#ggplot(redData, aes(x=alcohol, y=sulphates, color=quality, shape=quality)) +
#   geom_point() + 
#  geom_smooth(method=lm, se=FALSE, fullrange=TRUE)
#```






### White Wine Dataset ###

```{r}

cex.before <- par("cex")
par(cex = 0.7)
corrplot(as.matrix(cor(subset(whiteRawData, select=-c(color)))), cl.lim = c(-100, 100), type = "upper", insig = "blank", method = "color",
    addCoef.col="grey", 
    order = "AOE", tl.cex = 1/par("cex"),
    cl.cex = 1/par("cex"), addCoefasPercent = TRUE)
par(cex = cex.before)

```


Quality is correlated positively with Alcohol, negatively with Density and Chloride.

Information Gain shows Alcohol is an important attribute to determine quality of a wine in both of the datasets.


```{r}
attrEval(quality ~ ., data=select(whiteData, -color),  estimator = "InfGain") %>% 
        sort(decreasing = TRUE) %>% kable 
```


```{r}
set.seed(1024)
samples <- sample(1:nrow(whiteData), nrow(whiteData)*0.8)
train <- whiteData[samples, ]
test <- whiteData[-samples, ]
```



```{r}
fit <- rpart(quality ~ ., data = train[,c("quality","alcohol", "density", "chlorides")], method = 'class')

predicted <- predict(fit, test, type = "class")
conf_matrix <- confusionMatrix(predicted, test$quality)

conf_matrix$table
conf_matrix$overall["Accuracy"]


rpart.plot(fit)





























# References #

[Covariance and Correlation](https://towardsdatascience.com/let-us-understand-the-correlation-matrix-and-covariance-matrix-d42e6b643c22)



